Bayesian Inference Using Sequential Monte-Carlo Algorithm for Dynamic System Models

### 2020-08-09 MSc-dissertation.tex
New structure

\begin{titlepage}
    \vspace*{2in}
    % an acknowledgements section is completely optional but if you decide
    % not to include it you should still include an empty {titlepage}
    % environment as this initialises things like section and page numbering.
    \section*{Acknowledgements}

    This template is a slightly modified version of the one developed by
    Prof. Charles Duncan for MSc students in the Dept. of Meteorology. His
    acknowledgement follows:

    {\em This template has been produced with help from many former students who
    have shown different ways of doing things. Please make suggestions for
    further improvements.}

\end{titlepage}

@@

\chapter{Introduction}

 % This should contain a description of your project and the problem you
 % are trying to solve. Where appropriate you should also include
 % references to work which has already been done on your topic and
 % anything else which lets you set your work in context.

 % One of the things you will need to do is to ensure that you have a
 % suitable list of references.  To do this you should see \cite{ref:lam}
 % or some other suitable reference.  Note the format of the citation used
 % here is the style favoured in this department.  Here is another
 % reference \cite{ref:bloggs} for good measure.

 [background and motivation]

 [what is done]

While we are interested in the tissue regeneration among different species, mathematical models and techniques can help us to resolve complex interactions of cells and chemicals in a biological system. The objective process in this dissertation is an example of complex biology system found in tissue regeneration. Zebrafish has the ability to regenerate its spinal cord after injury; at the injury-site various immune cells are presented and complex interactions are found while they are regulating inflammation and repair.

Our interest lies in the modeling of the regeneration process that happens in the injury-site, especially in the parameter estimation and comparison of different models. According to available researches \cite{ref:Tsarouchas}, a mathematical model is proposed but the addressed interactions remains to be precisely defined, and quantitatively examined in experiments. However, we would like to explore the the `best fit' if the model which comes with a set of parameter estimates given the observed data. Also, we wish to compare different it with different models and possibly find the best model to describe the data without losing biological and mathematical properties in the regeneration process.

Fo the parameter estimation tasks, Bayesian inference techniques can help in the problems of over-fitting and local optima \cite{ref:abcsysbio}, compared to optimisation based methods. Moreover, approximate inference methods (Approximate Bayesian Computation, ABC) make likelihood-free inference possible, which is suitable for our task as writing down likelihood as a function expression can be hard.

Such methods introduce computation-intensive sampling-rejection pattern, where massive number of samples are drawn, followed by computations. As the sampling process is independent from each other in a certain period, the parallel execution becomes feasible and is expected to accelerated the parameter inference in a large scale.

This dissertation includes the implementations and experiments of parameter estimation using ABC SMC (sequential Monte-Carlo) method, and discussions on the results. Model comparison results are also included using the same method. Additionally, a performance test illustrates the strong-scaling performance of the algorithm.

\chapter{Background}

\section{Zebrafish spinal cord regeneration}

 [describe the process and how cells and cytokines are involved]

Compared to mammals, zebrafish can regenerate spinal cord after injury, and possibly recover swimming function. They are widely used in regenerative ability researches. Experiments found that although being cut off, the axonal of nerve cells can regenerate and bridge across the lesion site, which critically determined whether zebrafish can regain the swimming function \cite{axonal}. Tsarouchas et al. \cite{ref:Tsarouchas} studied the the complex interactions happen at the lesion site. The inflammation affects the regeneration, as if it is promoted, it accelerates the axonal regeneration; if it is inhibited, it reduces the regeneration. Further experiments showed that two kinds of cytokines, il-1$\beta$ and tnf-$\alpha$, affect the inflammation, which are dynamically controlled by immunes cells, i.e. cytokines works as media to control the regeneration. Here cytokines are small proteins, work as messenger to pass signal across different cells, which are similar to hormones and neurotransmitter.

An interaction map can be drawn, according to the evidences found in \cite{ref:Tsarouchas}. After injury, a massive immune response was observed, with the increasing numbers of several kinds of cells: macrophage, neutrophil and keratinocyte. The dynamical system consists of complex interactions between cells and cytokines, e.g. il-1$\beta$ promotes macrophages' increase. and as a major source, more macrophages leads to more expression of tnf-$\alpha$ mRNA. Consequently, tnf-$\alpha$ and il-1$\beta$ plays determinative roles in the axonal regeneration. Generally, il-1$\beta$ promotes the regeneration at early stages, but later has an inhibition effect, and tnf-$\alpha$ promotes the regeneration all the time. The presence if the cytokines also affects the number of cell, while as sources number of cells also affects cytokines' expression. Interactions between different kinds of cells are also considered in the system.

This is a dynamic system suitable for mathematical modelling, as a set of equations can be written to represent the dynamics as possible. Our interest lies in the modelling of the regeneration period of the dynamic system, and we wish to find that if we can quantitatively study the interactions iniside the system, based on the proposed hypothesis \cite{ref:Tsarouchas}, and our own hypothesis which may include different interactions.

The biological background of works in this dissertation is mostly related the the experimental evidences, quantitative data and proposed hypothesis in \cite{ref:Tsarouchas}. The published data on the quantitative measurement of cells and cytokines make the modelling and further analysis possible. Some other interactions proposed in our modelling stages are enlightened from Anderson et al. \cite{Anderson}.

\section{Mathematical modelling}

 [how this can be modelled as dynamic systems]

 [general topics of mathematical models: how to build model according to interactions; how to calibrate/ parameterise the model; how to solve the model; dynamics; how to evaluate the model]

Mathematical models can express real-life questions in mathematical forms, and provide qualitative or quantitative inside-look of the question through various mathematical tools and techniques. Mathematical models can be used for prediction, classification, testing our understanding or hypothesis and many other tasks, thus they are widely used to solve practical problems. However, for a model that describe the zebrafish spinal cord regeneration process, we were less interested to the model's predictive functionality; the main focus is on the explore of uncertainty in parameter estimates and comparison with alternative models. In this case, we wish to use modelling techniques to test our inference method, our hypothesis and hopefully an well-fitted models can be found and compared.

According to our understanding of the biological mechanism in the regeneration, hypothesis should firstly proposed, in a way which can represent the real  mechanism as much as possible. Usually a real-life problem, e.g. the spinal cord regeneration, consists of complex dynamics and interactions with many other environmental factors to be considered, An elaborate description of the real cases using a mathematical model sometimes can be impossible; assumptions and hypothesis are proposed to make necessary simplification and abstract for a final reasonably detailed model. Some qualitative models are simple but illustrative for the real-life observed patterns, e.g. Lotka-Volterra (L-V) equation for predator and prey models; they are usually analytically solvable and some properties are used to make predictions (e.g. steady states in ecosystem). If a more precise model is desired, usually more effects or interactions are considered and added as terms to a re-constructed model, e.g. environmental factors, a changing environmental capacity and predation from other species.

Models like L-V equation are mostly analytically solvable when analysing properties of the model, however, analytical solution for many models are not feasible, because of more complex dynamics, or the model itself is not a deterministic model. In the areas that are not analytically feasible, usually computational techniques are applied for approximate solutions. Solvers for different models are used to simulate the system or find roots for some spacial points.

For the regeneration process that we wanted to model, differential equations are our preference. Non-linear ordinary differential equations are used in this dissertation, where the only independent variable is time $t$ denoting the post-lesion time (hours), but further models can also consider other models e.g. stochastic models and partial differential equations. For ordinary differential equations (ODE) system, there are several available computational solvers using different algorithms, implemented in MATLAB\footnote[1] {\url{https://uk.mathworks.com/help/matlab/math/choose-an-ode-solver.html}}, Python\footnote[2]{\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.integrate.solve_ivp.html}} and many other languages. These solvers help to simulate the trajectories of the dependent variables, given initial values and a set of fixed parameters. The goodness of `fit' then can be measured by compared the simulated data and observed real data.

Often further researches on the modelled problems require the model to be deterministic on its parameters. Parameters in the dynamic system models usually have their practical significance, e.g. describing reaction speed, environmental capacity or the birth rate. There are many approaches to derive these parameters from data or experience. For a specific problem and a widely used models, the parameters' values can be read from previous literatures; however in many others cases, we need to find parameter's values specifically for our model and problem, where numerical values for parameters are more suitable for practical analysis using ODE solvers or other tasks.

To find the best parameters' values given observed data, either optimisation-based approaches (e.g. least-square (LS) fit) or inference approaches can be used. Optimisation-based approaches may face two major drawbacks: over-fitting and local optima \cite{ref:abcsysbio}. Bayesian inference approaches gain popularity in recent years, which can relief the concerns in over-fitting and provide a measurement for the uncertainty in the inferred values. Local optima can still be a major concern, and many efforts have be proposed for a more through explore of the parameter space.

\section{Bayesian inference}

 [how to parameterise the model]

 [how to infer the parameter given data]

 [likelihood-free inference and information about ABC SMC]

 [model comparison, ABC SMC for model comparison]

 Bayesian inference approaches put Bayes Rule in the centre:

 \begin{align}
    \label{eq:bayes}
    p(\theta|D) = \frac{p(D|\theta)p(\theta)}{P(D)}
\end{align}

where $p$ is the probability density function, $\theta$ denotes the parameter set, $D$ is the given observed data.

Alternatively, as $p(D)$ is fixed once the observed data is given, the rule can be rewritten as

\begin{align}
    \label{eq:bayes2}
    p(\theta|D) \propto l(D|\theta)p(\theta)
\end{align}

where $l$ represents the likelihood function. In words, Eqn \ref{eq:bayes2} can be described as `posterior is proportional to the product of likelihood and prior', and here posterior of parameters is our target. Prior $p(\theta)$ should express our initial believes or estimates on the parameter valeus before observing the data and conducting experiments, and likelihood function $l(D|\theta)$ is used to describe how well the parameter set $\theta$ explains the data $D$, in probability numbers. Once the data is given, $\theta$ is the only parameter in the likelihood function. Some optimisation-based methods try to find an estimation of parameter set $\theta$ s.t. maximise the likelihood, such methods will struggle when facing a problem where it is hard to specify an analytical expression of likelihood function, let alone the derivative.

Based on Eqn \ref{eq:bayes2}, to find the posterior of parameter set, prior can be set according to the problem context and our understanding and cover a reasonable ranges; computational methods can be used to approximate likelihood function, e.g. Markov Chain Monte Carlo (MCMC) \cite{ref:MCMC}. As another alterative solution, Approximate Bayesian computation (ABC) approaches do not focus on the evaluation of likelihood, but focused on simulating the data from the given model and observed data, then used the simulated data to approximate the desired posterior. ABC-rejection \cite{ABC_rejection} is a simple example of the ABC process, and further algorithms e.g. sequential Monte-Carlo (ABC SMC)\cite{Toni} are proposed to improve the efficiency.

General steps for ABC algorithm includes the following (ABC-rejection):

\begin{enumerate}
    \item Draw a sample $\theta^*$ from parameters' prior distribution. $\theta^*$ is a vector that consists values for each parameter
    \item Plug $\theta^*$ into the model, and use the model to generate simulated data $D^*$, where $D^*$ is of the same form as the observed data $D$, e.g. includes same summary statistics
    \item The discrepancy between the simulated data and observed data is measured by a distance function $d(D^*, D)$. For a pre-fixed threshold value $\epsilon$, if $d(D^*, D)<\epsilon$ then the sample $\theta^*$ is accepted
    \item Repeat 1, 2 and 3 until enough numbers of samples are accepted. The posterior distribution can be approximated by these samples
\end{enumerate}

In our study of the regeneration model, full trajectories of the dependent variables are used as the data to be compared, as our goal is to find a model and a parameter set that can intuitively recover the real process as much as possible. The mean values at each time point for each dependent variable is calculated to form $D$, and for each sample, the ODE is accumulatively solved at the same time points and generate the simulated trajectories $D^*$. Using other statistics to summary the raw data are also possible bur care must be taken as the choice of summary statistics can largely affect the resultant posterior \cite{summaryD, summaryD2}.

Instead of fixing the threshold value, the algorithm can be more adaptive by using a sequence of deceasing threshold values $\epsilon_t$, and derive a new population (called generation) of accepted sampled from the previous accepted samples (called particles). An example is ABC SMC \cite{Toni}, which can significantly reduce the required samples compared to ABC-rejection \cite{ref:disease}:

\begin{enumerate}
    \item 
\end{enumerate}

\section{Software tools}

 [existing tools and software for this task]

 [workflow and developing process]

\chapter{Mathematical modelling}

Mathematical models can describe different kinds of dynamic systems, and thus can be used as a guide for prediction and analysis. An ideal model in our case, can represent reasonable interactions/effects between cells and cytokines and recover the observed  data features in different time point.

Proposed models for the regeneration process are in the form of ordinary differential equations (ODE), specifically the time differential form. Terms in the ODE are mostly explainable and corresponds interaction paths.

\section{Observed data}

Our models is built on the basis of the existing experiments data form Tsarouchas et al.\cite{ref:Tsarouchas}. The measurement data include the number of three kinds of cells (neutrophil $N$, macrophage $\Phi$ and microglia) and the relative concentration of four cytokines (il-1$\beta$, tnf-$\alpha$, tgf-$\beta$1a and tgf-$\beta$3). As proposed in \cite{ref:Tsarouchas}, neutrophil and macrophage play important roles in the promotion of spinal cord regeneration with il-1$\beta$ and tnf-$\alpha$ being the mediation. According to this, our current models focus on the changes of four variables $N$, $\Phi $, $\beta$ (for il-1$\beta$) and $\alpha$ (for tnf-$\alpha$). $N$ and $\Phi$ is of the unit `number of cells', $\beta$ and $\alpha$ is of the unit `relative concentration' (compared to the value at first time point).

It is noted that the variance of the measured data is relatively high. The summary statistic used for the parameter estimations is mean of measurement data at each time point, assuming that measured data is Gaussian-like distributed. To validate this, the distribution of the measured data points is plotted and examined to see if the mean value can represent the distribution. The result is that at most time points the measurement values are Gaussian-distributed, although some distributions are skewed. One abnormal distribution is observed at time point 120 h post-lesion (hpl) for macrophage where there are two concentrations. Mean value can summarise most data and thus is still used as the target observed data. A plot of the mean of the four variables is shown in Figure \ref{fig:obs_data}.

\begin{figure}
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/obs_curve.png}}
    \end{center}

    \caption[Mean of the observed data]%
    {Mean of the observed data for neutrophil ($N$), macrophage ($\Phi$), il-1$\beta$ and tnf-$\alpha$, from experiment results of \cite{ref:Tsarouchas}. Error bars indicate standard error of mean}
    \label{fig:obs_data}

\end{figure}

\section{Hypothesis and Models}

5 models in total are proposed according to different hypothesis. At first our tests and implementations of ABC SMC for parameter estimations use only the basic model for developing propose. After the parameter estimation framework is built and tested, more models are proposed, in order to calibrate and adjust the basic model such that it can represent the observed regeneration process better or can be used to test our hypothesis.

All these models assume the involved interactions is within two kinds of cells (neutrophil and macrophage) and two kinds of cytokines (il-1$\beta$ and tnf-$\alpha$) and use the data presented in Figure \ref{fig:obs_data} for the inference task. Interactions or effects from other cells or cytokines is not considered as there might not be corresponding data available.

\subsection{Basic model}

A preliminary model is proposed according to \cite{ref:Tsarouchas} and used to build and test the inference framework code. An interaction map of the model is shown in Figure \ref{fig:m1}. This model is a simplification a the process described in \cite{ref:Tsarouchas} with some minor interactions ignored. To describe the parameters' units, we denote the unit of $N$ and $\Phi$ i.e. number of cells as `cell', and denote the unit of $\beta$ and $\alpha$ i.e. relative mRNA expression as `unit' for simplicity.

\begin{figure}
    \begin{center}
        \resizebox{0.4\hsize}{!}{\includegraphics{fig/model1.png}}
    \end{center}

    \caption[Interactions modelled in the basic model]%
    {Interactions modelled in the basic model (model 1) based on Tsarouchas et al.\cite{ref:Tsarouchas}. Lines ended with arrow represent promoting effect, lines ended with T-connectors represent inhibition}
    \label{fig:m1}

\end{figure}

It is assumed that there are negative feedbacks for all the variables. MORE DESCRIBE.

\begin{align}
    \label{eq:model1}
    \begin{split}
        &\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_N+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\
        &\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\lambda_\Phi+\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\
        &\frac{\mathrm{d} \beta}{\mathrm{d} t}=\frac{s_{\beta N}N}{1+i_{\beta\Phi}\Phi}-\mu_\beta\beta\\
        &\frac{\mathrm{d} \alpha}{\mathrm{d} t}=s_{\alpha\Phi}\Phi-\mu_\alpha\alpha
    \end{split}
\end{align}

\begin{table}[h!]
    \centering
    \begin{tabular}{|c c c|}
        \hline
        Parameter            & Definition                                       & Units                    \\ [0.5ex]
        \hline\hline
        $\lambda_N$          & Self-increase rate of neutrophil                 & $cell/h$                 \\
        $\kappa_{N\beta}$    & Promoting effect coefficient by il-1$\beta$      & $cell/(unit\cdotp h)$    \\
        $\mu_N$              & Coefficient of negative feedback of $N$          & $h^{-1}$                 \\
        $\nu_{N\Phi}$        & Coefficient of inhibition of both $N$ and $\Phi$ & $cell^{-1}\cdotp h^{-1}$ \\
        \hline
        $\lambda_\Phi$       & Self-increase rate of macrophage                 & $cell/h$                 \\
        $\kappa_{\Phi\beta}$ & Promoting effect coefficient by il-1$\beta$      & $cell/(unit\cdotp h)$    \\
        $\mu_\Phi$           & Coefficient of negative feedback of $\Phi$       & $h^{-1}$                 \\
        \hline
        $s_{\beta N}$        & Production rate from $N$                         & $unit/(cell\cdotp h)$    \\
        $i_{\beta\Phi}$      & Coefficient of inhibition to the production      & $cell^{-1}$              \\
        $\mu_\beta$          & Coefficient of negative feedback of $\beta$      & $h^{-1}$                 \\
        \hline
        $s_{\alpha\Phi}$     & Production rate from $\Phi$                      & $unit/(cell\cdotp h)$    \\
        $\mu_\alpha$         & Coefficient of negative feedback of $\alpha$     & $h^{-1}$                 \\
        \hline
    \end{tabular}
    \caption{Parameters introduced in the basic model (model 1) [REMOVE unit]}
    \label{table:m1}
\end{table}

\subsection{Alternative models}

\paragraph{Model 2 and model 3}

As the observed data indicate, this dynamic system has a steady state where the  inflammation is resolved and immune cells should not be present at the injury site. Regarding this, the self-increase parameter $\lambda$ cannot be constant, thus a exponentially decaying $\lambda$ term is introduced and model 2 is proposed. (Eqn. \ref{eq:model2}). Also, the inhibition of il-1$\beta$ production, i.e. term $i_{\beta\Phi}\Phi$ is considered to be ignored for a more simple model, in which case the relative expression of il-1$\beta$ is only affected by the number of neutrophil and the negative feedback from itself. This case corresponds to model 3, written as Eqn. \ref{eq:model3}.

Model 2 and model 3 introduce one extra parameter $a$, which is a coefficient in the exponentially decaying $\lambda$, determining the decay speed, with the unit $h^{-1}$.

\begin{align}
    \label{eq:model2}
    \begin{split}
        &\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_Ne^{-at}+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\
        &\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\
        &\frac{\mathrm{d} \beta}{\mathrm{d} t}=\frac{s_{\beta N}N}{1+i_{\beta\Phi}\Phi}-\mu_\beta\beta\\
        &\frac{\mathrm{d} \alpha}{\mathrm{d} t}=s_{\alpha\Phi}\Phi-\mu_\alpha\alpha
    \end{split}
\end{align}

\begin{align}
    \label{eq:model3}
    \begin{split}
        &\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_Ne^{-at}+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\
        &\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\
        &\frac{\mathrm{d} \beta}{\mathrm{d} t}=s_{\beta N}N-\mu_\beta\beta\\
        &\frac{\mathrm{d} \alpha}{\mathrm{d} t}=s_{\alpha\Phi}\Phi-\mu_\alpha\alpha
    \end{split}
\end{align}

Model 3 can be regarded as a simplification of model 2, as it can be treated as model 2 with parameter $i_{\beta\Phi}=0$. Among the proposed three models, model 1 is a naive one that was proposed at very first time and used as an ODE `template' to build and test parameter inference framework. As the implementation was successful, model 2 and 3 was proposed as we were trying to calibrate some terms and find a better model. After fitting, model 2 is supposed to be better than model 1 as it corrects the problem that appears at the final time points (which relate to the steady state). Model 3 makes a small simplification on model 2 and is theoretically less `general' than model 2.

\paragraph{Model 4 and model 5}

After the first model selection experiment, it was found that some significant features presented in the observed data were not recovered by any of the existing model. To resolve this, attempts were tried to introduce more interactions within the dynamic system considering the biological and  mathematical context. Extra promoting effect to the expression of tnf-$\alpha$ was considered, by either adding a phenomenological term (which means the same effect as directly promoting but the underlying mechanism is unclean) or adding a term that represents a promoting effect to the production process of tnf-$\alpha$, namely model 4 (Eqn. \ref{eq:model4}) and model 5 (Eqn. \ref{eq:model5}).

\begin{align}
    \label{eq:model4}
    \begin{split}
        &\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_Ne^{-at}+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\
        &\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\
        &\frac{\mathrm{d} \beta}{\mathrm{d} t}=s_{\beta N}N-\mu_\beta\beta\\
        &\frac{\mathrm{d} \alpha}{\mathrm{d} t}=s_{\alpha\Phi}\Phi-\mu_\alpha\alpha+d_{\beta\alpha}\beta
    \end{split}
\end{align}

\begin{align}
    \label{eq:model5}
    \begin{split}
        &\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_Ne^{-at}+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\
        &\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\
        &\frac{\mathrm{d} \beta}{\mathrm{d} t}=s_{\beta N}N-\mu_\beta\beta\\
        &\frac{\mathrm{d} \alpha}{\mathrm{d} t}=(s_{\alpha\Phi}+f_{\beta\alpha}\beta)\Phi-\mu_\alpha\alpha
    \end{split}
\end{align}

\begin{table}[h!]
    \centering
    \begin{tabular}{|c c c|}
        \hline
        Parameter         & Definition                                                     & Units                    \\ [0.5ex]
        \hline\hline
        $d_{\beta\alpha}$ & Coefficient of promoting effect from $\beta$                   & $h^{-1}$                 \\
        \hline
        $f_{\beta\alpha}$ & Coefficient of promoting the production of $\alpha$ by $\beta$ & $cell^{-1}\cdotp h^{-1}$ \\
        \hline
    \end{tabular}
    \caption{Parameters introduced in model 4 and 5}
    \label{table:m45}
\end{table}

\subsection{Model evaluation and comparison}

[general topics on evaluating a model]

[bayes factor for model selection]

\subsection{Limitations}

[hypothesis]

[available data]

[model misspecification]

% \begin{figure}

% \begin{center}
% \resizebox{0.30\hsize}{!}{\includegraphics{logos/crest_bw}}
% \end{center}

% \caption{The University Crest}
% \label{fig:eucrest}

% \end{figure}

% see the man page for dvips for details of the special command which is
% much more powerful than is shown here. It allows offsets in the
% horizontal and vertical and scaling in x and y.

% choosing suitable values for offset and scale can be a tiring matter
% of trial and error.

% note that labels do not need to include a description of the object
% they are labelling but it can be helpful, eg \label{fig:figurename}.

% You can use a label on a figure to refer to it later. The university
% crest is in \ref{fig:eucrest}. Note that you should not use phrases like
% ``the figure above'' or ``the following figure'' since Latex may move
% the figure relative to the text if it cannot be fitted onto the current page.

\chapter{Parameter estimation and model comparison}

Estimation of  parameters in the proposed models is the main task of this section, which involves the implementation of ABC SMC and several experiments. ABC SMC-based model selection is another task that can be easily conducted with the developed parameter inference framework.

The build and test of code were under \verb|Python| environment with \verb|pyABC|\cite{ref:pyabc}. Some other code e.g. shell script and \verb|R| code are also used to perform the experiments and analyse the results. Software and system environment used in our implementation can be found at Appendix A. The build and test were performed on local computers and HPC facilities available within EPCC.

\section{Implementations and code}

 [ABC implementations details, e.g. distance, population, ODE related functions]

 [how the code is developed and built]

According to our preliminary studies, ABC SMC framework implemented using \verb|pyABC| in \verb|Python| was adopted in this project. \verb|pyABC| is a popular SMC software packages \cite{ref:pyabc} that has been used form many SMC studies and applications \cite{ref:adpt_pop}. \verb|pyABC| provides an open-source framework for likelihood-free inference, which is an implementation of ABC SMC algorithm and a tool-box for our own inference and analysis tasks. Besides, it supports multi-core execution implemented using \verb|multiprocessing| package for scaling-up, which is of our interests and suitable for the followed performance experiments.

The inference framework code for the project was developed and tested in local environment (macOS 10.15.6, see Table \ref{table:local_macine}) at first, then the functional version was deployed on compute node of ARCHER and Cirrus for parameter inference experiments and model selection task. The scaling-up experiment was performed on Cirrus, and the profiling and its analysis were performed on local machine. Some development and debug were on the remote machines to ensure experiments run correctly on compute node.

The ODE model-related code and utilities were firstly developed to enable reusable functions, variables and data calls. It includes the code for the ODE models, ODE solver, data structures, data format transform functions etc. By doing this in a separate code file, we can change options and activate ABC SMC easily in the implementation code file, without creating definition and duplicated code for the models and parameter priors.

Some modifications were made to the source code of \verb|pyABC|. For example, the laboratorial measured data were not considered `complete' for modeling and analysis purpose: the available time points for cells and cytokines are different, e.g. cell number was not measured at 0.25 hpl and cytokines' expression was not measured at time point 120 hpl. These missing values were treated as `ignored' when calculating the distance between observed data and simulated data. To cope with this built-in distance function code is modified.

Analysis code was also wrapped up into functions which can be easily called after an ABC SMC run is finished, which includes multiple visualisations and summary statistic of the results. For different tasks there are also separated analysis code file. Other analysis tools used in the project include built-in database visualisation tool, R code and Microsoft Excel.

The project code was managed via \verb|git| version-control and available as a GitHub repository\footnote{\url{https://github.com/chaolinhan/MSc_Project}}; it also made the cross-hardware synchronisation of code versions convenient.

After several development iterations, code files for ABC SMC implementation now mainly had two types: code files for infer-back experiments (udsing synthetic data), and code files for parameter inference (using experimental data). Model selection can be easily done by adding more model to the implementation code file.

% \section{Parameter estimation}

% [implementation details and options/ settings that are affecting the ABC]

\section{ABC SMC and model hyperparameters experiments}

As the key focus of this project is on the parameter estimations of dynamical system, the ability of inference on the proposed models was firstly examined before actual inference using the experimental observed data.

In the first part, synthetic data is used with known true parameter values. The algorithm implementation is evaluated by the efficiency and goodness of resultant model under different implementation options and ABC SMC hyperparameters.

Then according to findings in the first part, several options with good performance are tried in the inference using experimental observed data (Figure \ref{fig:obs_data}). To obtain a more accurate and general results, these experiments are repeated for 3 or more times.

    [including the options that are tested/ tried/ not adopted]

    [including experiments results and discussions]

    [how do these options correspond to biological model and how to set them accordingly for proposed model]

The ABC SMC inference can have largely different performance when using different hyperparameters and implementation options. To firstly test the ability of inference and observe the results of different options, synthetic data with known true parameter values are used as target data. The true value of parameters that are used to generate synthetic data is listed in Appendix B.1. These true values are firstly obtained by fitting a least square fit of model 1 onto the experimental data (Figure \ref{fig:infer_back_data}), the target data is then generated using these parameter values via ODE solver in \verb|Python|.

\begin{figure}[h!]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/LS.png}}
    \end{center}

    \caption[Synthetic data generated with known parameter values]%
    {Synthetic data generated with known parameter values (line plot), compared to experimental data (scatter plot). Known parameter values are obtained from a least square fitting of the observed data}
    \label{fig:infer_back_data}

\end{figure}

[TABLE for known parameter values]

[why use synthetic data]

Among them the following topics are studied using the synthetic data.

\subsection{Perturbation kernels}

Perturbation kernels work in the sampling process. In each generation $t$, samples are firstly taken from the previous population $\{\theta_{t-1}\}$ with weights $\{w_{t-1}\}$, then perturbed using the perturbation kernel $K(\theta|\theta^*)$. We keep sampling until $N$ particles are accepted, where $N$ is the pre-set population size. After that the new weights are calculated and normalised.

\begin{figure}[t!]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/smc.png}}
    \end{center}

    \caption[ABC SMC sampling process]%
    {ABC SMC sampling process. $\theta^*$ denotes a sample drawn from previous population, which is used to generate simulated data $D^*$. $d$ is a distance measurement, $d(D^*,D)$ represents the discrepancy between observed data and simulated data. for each population $t$, $\epsilon_t$ is a threshold criterion to determine whether to accept that drawn sample}
    \label{fig:smc}

\end{figure}

The perturbation kernel is called in the sampling of every particles and determines how the new perturbed particle is chosen, thus it is influential to both computation complexity and the resultant posterior distribution. Generally, a local perturbation kernel may face the risk of being stuck in local modes (e,g., local optimal), but it may need less computational operations, or could generate a population with a higher acceptance rate if the successive epsilon values are close; A kernel with wide variance, or spreading out in a large space could help in resolving the local optimal problem by a more thoroughly exploration of the parameter space, however it can be more computation-intensive and result in a lower acceptance rate. A desired optimal kernel should balance the trade-offs; their property and criteria is discussed in \cite{ref:kernel}.

There are several common choice of perturbation kernels. Among them multivariate normal kernel and local M-nearest neighbour model is preferred to be applied on our models. A covariance matrix $\Sigma_t$ of accepted particles is calculated form previous generation and used in multivariate normal kernel: $K(\theta|\theta_{t-1})\sim\mathcal{N}(\theta_{t-1}, \Sigma_t)$. It is illustrated to be more efficient than uniform kernel and component-wise normal kernel in relecting the true posterior structure \cite{ref:kernel}. It has been proved to perform well in several dynamic system models \cite{ref:abcsysbio, ref:compare, ref:disease} for the parameter estimation and model selection tasks. The $scaling\in(0,1]$ parameter in \verb|pyABC| will be multiplied to the covariance to produce a `narrower distributed' perturbation result.

Local M-NN kernel provided by \verb|pyABC| is also tried to provide a comparison. A local kernel density estimation (KDE) fit is used with M nearest neighbours considered.

The kernel experiments is designed to explore the efficiency of SMC on our dynamical systems. Given the same fixed threshold schedule, kernels that need less total samples and have higher acceptance rate would be our preference. The experiments compared multivariate normal kernel with local MNN kernels with different parameters, using synthetic data to infer back the parameters. The acceptance rate among each time point and total required samples are compared after the experiment to give suggestions on the kernel selection in the real data inference. As the threshold schedule is fixed, the final population should have similar discrepancy to the target data thus here the goodness of fit i.e. recover of target data trends/features and errors of the inferred parameters compared to true values is not discussed.

\subsubsection{Kernel experiment results}

To compare kernels, a fixed schedule schedule is used with minimal epsilon i.e. $\epsilon_t$ set to 10.0. From the total required samples graph Figure \ref{fig:kernel1}, the efficiency can be compared. Among the tested kernels, the local M-NN with M=50 has the best performance: it requires the least number of samples and
has higher acceptance rates among almost all generations. Local M-NN with M=750 is the slowest kernel.

For local M-NN kernels, generally a greater M will lead to lower acceptance rate and
more required particles in each generation. As shown in Figure \ref{fig:acceptance1}, local M-NN with M=750 is the lowest curve. Consequently, if greater M is test, then it will have even lower acceptance rates; the maximum M=2000 (whole population is considered) will have the lowest acceptance rates.

\begin{figure}
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/kernel1.png}}
    \end{center}

    \caption[Total required samples of different kernels]%
    {Total required samples of different kernels after 20 populations (2000 particles in each population). Different color represents different generations (bottom to top: population 1 to population 20)}
    \label{fig:kernel1}

    \vspace*{\floatsep}

    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/acceptance1.png}}
    \end{center}

    \caption[Acceptance rates of different kernels]%
    {Acceptance rates of different kernels in 20 populations. Each population has 2000 particles}
    \label{fig:acceptance1}

\end{figure}

Multivariate normal kernel has a performance between local M-NN M=250 and M=100. It proves that rather than a trivial normal kernel, multivariate normal kernels is more efficient facing the concentrations of joint distributions among multiple parameters \cite{ref:kernel}. `Scaling' option can narrower the distribution calculated from the last population, thus making the kernel more `local' by sampling under a smaller variance. Similar to small M in local M-NN, small `scaling' is also more efficient in our experiment of the model 1. Besides the fixed threshold schedule, a median threshold schedule with the same final threshold $\epsilon_{20}=10.0$ is also tested and similar results are observed (Figure \ref{fig:kernel2} and \ref{fig:acceptance2}, Appendix B2).

>>>>

% \begin{titlepage}
%     \vspace*{2in}
%     % an acknowledgements section is completely optional but if you decide
%     % not to include it you should still include an empty {titlepage}
%     % environment as this initialises things like section and page numbering.
%     \section*{Acknowledgements}

%     This template is a slightly modified version of the one developed by
%     Prof. Charles Duncan for MSc students in the Dept. of Meteorology. His
%     acknowledgement follows:

%     {\em This template has been produced with help from many former students who
%     have shown different ways of doing things. Please make suggestions for
%     further improvements.}

% \end{titlepage}

@@

\input{introduction/introduction.tex}

\input{background/background.tex}

\input{modelling/modelling.tex}

\input{implementation/implementation.tex}

\input{performance/performance.tex}

\input{conclusion/future_works.tex}

\input{conclusion/conclusion.tex}
